---
title: "PCA for Image Compression"
editor_options: 
  chunk_output_type: inline
format:
  html:
    toc: true
    toc-title: Contents
    toc-location: left
---

```{r}
#| include: false
library(tidyverse)
library(reshape2)
library(patchwork)

# https://www.pmassicotte.com/posts/2022-08-15-removing-whitespace-around-figures-quarto/
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
```

```{r}
#| echo: false

## cite https://www.geeksforgeeks.org/r-language/how-to-read-many-files-in-r-with-loop/
##https://quarto.org/docs/output-formats/html-basics.html

img_filenames = list.files(path = "images", pattern = "\\.csv$")
images = list()
for (file in img_filenames) {
  filepath <- paste0("images/", file)
  # read in image matrix
  img_mx <- read.csv(filepath) |> data.matrix()
  images[[file]] <- img_mx
}
```

```{r}
#| echo: false
get_plot_obj <- function(mat, label = ""){
  df <- as.data.frame(as.table(mat)) 
  plot_obj <- ggplot(df, aes(Var1, Var2, fill = Freq))+
    ggtitle(label)+
    geom_tile(show.legend = FALSE) +
    scale_fill_gradient(low = "black", high = "white") +
    coord_equal()+
    theme_void()
  return(plot_obj)
}
```

```{r}
#| echo: false
first_evector_pc <- function(X, i =50){
  pca <- prcomp(X,  center=FALSE, scale.=FALSE)
  pc1 <- pca$x[,1:1]
  evec1 <- pca$rotation[,1:1]
  df_pc1 <- as.data.frame(as.table(pc1)) 
  df_evec1 <- as.data.frame(as.table(evec1))
  
  new <- pc1 %*% t(evec1)
  df_new <- as.data.frame(as.table(new)) 
  
clear_axes <- theme(
  axis.text = element_blank(),   
  axis.ticks = element_blank(),  
  axis.line = element_blank()  )  
  
  p1 <- ggplot(df_pc1, aes(Var1, 1, fill = Freq))+
    geom_tile(show.legend = FALSE) +
    scale_fill_gradient(low = "black", high = "white") +
    clear_axes+
    labs(x = "First Eigenvector (PC score weight)", y = "")+
    scale_x_discrete(position = 'top')+
    ylim(0, 2) +
    coord_fixed(ratio = 50)

  
  
  p2 <- ggplot(df_pc1, aes(1, Var1, fill = Freq))+
    geom_tile(show.legend = FALSE) +
    scale_fill_gradient(low = "black", high = "white") +
    labs(x = "", y ="First Principal Component Scores")+
    xlim(0, 2) +
    coord_fixed(ratio = 1/i)+
    clear_axes

  
  p3 <- ggplot(df_new, aes(Var1, Var2, fill = Freq))+
    geom_tile(show.legend = FALSE) +
    scale_fill_gradient(low = "black", high = "white") +
    coord_fixed()+
    theme_void()
  
 plot_spacer()+ p1 + p2 + p3+
    plot_layout()
}


```

```{r}
#| include: false
get_compressed <- function(X, k){
  pca <- prcomp(X,  center=FALSE, scale.=FALSE)
  new_mx <- (pca$x[,1:k] # PC scores for k columns
            %*% t(pca$rotation[,1:k])) # Transpose ofirst k cols of the matrix of variable loadings (i.e., a matrix whose columns contain the eigenvectors
  error <- norm((X-new_mx), type = "F")
  return(c(compressed_mat = mat, error = error))
}

get_images <- function(X, k_list){
  pca <- prcomp(X,  center=FALSE, scale.=FALSE)
  original <- get_plot_obj(X)
  approx_imgs <- list(original)
  for (k in k_list) {
    new_mx <- (pca$x[,1:k] # PC scores for k columns
            %*% t(pca$rotation[,1:k])) # Transpose ofirst k cols of the matrix of variable loadings (i.e., a matrix whose columns contain the eigenvectors
    label <- paste0("K = ", k)
    plot_obj <- get_plot_obj(new_mx)
    approx_imgs[[length(approx_imgs)+1]] <- plot_obj
  }
  return(approx_imgs)
}

get_error_plot <- function(X){
  pca <- prcomp(X,  center=FALSE, scale.=FALSE)
  k_list = 1:(ncol(pca$x))
  error_list <- c()
  #print("starting interation over k")
  for (k in k_list) {
    new_mx <- (pca$x[,1:k] # PC scores for k columns
            %*% t(pca$rotation[,1:k])) # Transpose ofirst k cols of the matrix of variable loadings (i.e., a matrix whose columns contain the eigenvectors
    error <- norm((X-new_mx), type = "F")
    error_list <- c(error_list, error)
  }
  df_error <- tibble(
    k = k_list,
    error = error_list)
  error_plot <- df_error |> 
    ggplot(aes(k, error))+
    geom_point(aes(size =2),show.legend = FALSE)+
    geom_line()+
    scale_y_continuous(limits = c(0, NA))
  return(error_plot)
}


```

## Image 1

::::: grid
::: g-col-8
The approximation is sufficient at K = 1

The white line that is the most distinctive feature of the image is clear in the principle component scores. Outside of that, the values seem to be random-appearing noise.

```{r}
#| echo: false
first_evector_pc(images$image1.csv,500)
```
:::

::: g-col-4
```{r}
#| echo: false
plot(get_error_plot(images$image1.csv))
```
:::
:::::

```{r}
#| echo: false
#| layout-ncol: 1
#| crop: true
#| fig-cap: 
#|   - "Original Image"
#|   - "K = 1"
#|   - "K = 2"
#|   - "K = 3"
#|   - "K = 4"
#|   - "K = 5"
#|   - "K = 6"
#|   - "K = 7"
#|   - "K = 8"
#|   - "K = 9"

imgs <- get_images(images$image1.csv, 1:9)
for (im in imgs){
  plot(im)
}
```

## Image 2

::::: grid
::: g-col-8
The approximation is sufficient at K = 1

The first eigenvector, weighting the pc scores, seems capture more information than the pc score itself, as it's inverse seems to correspond exactly to the vertical streaks in the image.

```{r}
#| echo: false
first_evector_pc(images$image2.csv,500)
```
:::

::: g-col-4
```{r}
#| echo: false
plot(get_error_plot(images$image2.csv))
```
:::
:::::

```{r}
#| echo: false
#| layout-ncol: 1
#| crop: true
#| fig-cap: 
#|   - "Original Image"
#|   - "K = 1"
#|   - "K = 2"
#|   - "K = 3"
#|   - "K = 4"
#|   - "K = 5"
#|   - "K = 6"
#|   - "K = 7"
#|   - "K = 8"
#|   - "K = 9"

imgs <- get_images(images$image2.csv, 1:9)
for (im in imgs){
  plot(im)
}
```

## Image 3

::::: grid
::: g-col-8
The dog can clearly be seen at k = 9, and it appears the decrease in error for each successive k begins to rapidly diminish around k = 25. For this reason, k = 25 may be a good approximation.

The first PC and eigenvector describe the rectangular central mass of the dog.

```{r}
#| echo: false
first_evector_pc(images$image3.csv,50)
```
:::

::: g-col-4
```{r}
#| echo: false
plot(get_error_plot(images$image3.csv))
```
:::
:::::

```{r}
#| echo: false
#| layout-ncol: 5
#| crop: true
#| fig-cap: 
#|   - "Original Image"
#|   - "K = 1"
#|   - "K = 2"
#|   - "K = 3"
#|   - "K = 4"
#|   - "K = 5"
#|   - "K = 6"
#|   - "K = 7"
#|   - "K = 8"
#|   - "K = 9"

imgs <- get_images(images$image3.csv, 1:9)
for (im in imgs){
  plot(im)
}
```

## Image 4

::::: grid
::: g-col-8
Even the original image is not super clear - an image around k = 15 may be a good approximation.  

The first PC and eigenvector primarily describe the white rectabgle in the upper right.

```{r}
#| echo: false
first_evector_pc(images$image4.csv,50)
```
:::

::: g-col-4
```{r}
#| echo: false
plot(get_error_plot(images$image4.csv))
```
:::
:::::

```{r}
#| echo: false
#| layout-ncol: 5
#| crop: true
#| fig-cap: 
#|   - "Original Image"
#|   - "K = 1"
#|   - "K = 2"
#|   - "K = 3"
#|   - "K = 4"
#|   - "K = 5"
#|   - "K = 6"
#|   - "K = 7"
#|   - "K = 8"
#|   - "K = 9"

imgs <- get_images(images$image4.csv, 1:9)
for (im in imgs){
  plot(im)
}
```
